Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-06-17 docs/paper.Rmd
write theoretical background (intergenerational q-q model)

Becker’s early work introduced an economic approach to fertility, treating children analogous to “durable goods” that yield utility to parents but come at a cost \citep{doepke2015gary}. In Becker’s 1960 model, a household derives satisfaction from the number of children ($n$) and from the quality of each child ($q$), alongside conventional consumption of other goods ($y$). A simple representation is a utility function:

@@

While the early Q–Q models were static (one-period) representations, subsequent contributions extended the framework to consider fertility and child investment over multiple periods or even multiple generations. A central development in this literature was the incorporation of intergenerational human capital dynamics, where parents derive utility not only from the number and quality of children in the present, but also from the long-run outcomes of their offspring. These extensions allowed child quality to evolve endogenously across time, rather than being determined solely within a single period.

>>>>

\subsubsection{Becker’s Static Model}
Becker’s early work introduced an economic approach to fertility, treating children analogous to “durable goods” that yield utility to parents but come at a cost \citep{doepke2015gary}. In Becker’s 1960 model, a household derives satisfaction from the number of children ($n$) and from the quality of each child ($q$), alongside conventional consumption of other goods ($y$). A simple representation is a utility function:

@@

\subsubsection{Intergenerational Models: Altruism and Child Endowments}

While the early Q–Q models were static (one-period) representations, subsequent contributions extended the framework to consider fertility and child investment over multiple periods or even multiple generations. The main development in this literature was the incorporation of intergenerational human capital dynamics, where parents derive utility not only from the number and quality of children in the present, but also from the long-run outcomes of their offspring. These extensions allowed child quality to evolve endogenously across time, rather than being determined solely within a single period.

One of the earliest and most influential models of this kind was proposed by \citet{becker1976child}. In their formulation, each child enters the world with an exogenous endowment $E$, which may reflect factors such as cognitive ability (e.g., measured IQ or language acquisition speed), early health status (e.g., birth weight or incidence of neonatal complications), genetic predispositions (e.g., risk for chronic illness, temperament, or neurodevelopmental traits), or family background characteristics (e.g., parental education, household stability, or neighborhood conditions). Furthermore, parents can augment this endowment by investing resources $q$ in the form of education, nutrition, and other quality-enhancing inputs. The effective adult human capital of the child might be expressed as $H = E + f(q)$ (in a simple additive form) or a multiplicative variant $H = E \cdot f(q)$, where $f(q)$ is an increasing concave function of parental investment.

\citet{becker1976child} emphasized that variation in endowments $E$ can shape how parents allocate investments $q$ across children. When the productivity of investment increases with endowment, parents may concentrate resources on children with higher $E$, who are more likely to convert additional investment into future success. In other cases, parents may attempt to compensate for lower endowments by directing greater investment toward disadvantaged children. Put simply, child-specific variation in initial conditions affects not only outcomes but also strategic parental choices. Because of this, the relationship between income and the demand for child quality is not uniform. The income elasticity of demand for quality may differ across households, depending on the distribution of endowments within the family and across the broader population.

Furthermore, \citet{becker1976child} showed that at low income levels, much of what constitutes child quality comes from exogenous endowments, i.e., factors like public education, neighborhood environment, or access to basic healthcare that are not privately purchased. In these settings, small increases in parental income may not lead to significant changes in fertility or investment behavior. Since most of the child’s future outcomes are determined by the fixed endowment component, marginal investment plays a smaller role. However, as income rises, private resources become a larger part of what determines quality, and the classic Q–Q trade-off begins to shape behavior. Parents begin to allocate more income toward fewer children in order to enhance quality through direct investment. Under certain theoretical conditions, such as equal utility elasticities for quantity and quality, this framework produces a non-monotonic relationship between income and fertility. Fertility may decline as income rises at first, which reflects the desire to invest more intensively per child. However, beyond some point, once the marginal return to investment begins to flatten or saturate relative to the fixed endowment, fertility may increase again.

The U-shaped prediction emerges only under specific assumptions, and its validity depends on both the shape of the utility function and how endowments relate to parental background. More broadly, \citet{becker1976child} enriched the Q–Q framework by incorporating elements that reflect real-world variation. They argued that not all differences in child outcomes are the result of deliberate parental choice. Random factors, biological traits, and socioeconomic settings play a role. In this light, public policies, such as subsidized schooling, early childhood programs, or universal healthcare, can influence private fertility and investment decisions by shifting the effective value of $E$ across the population. If government programs raise the floor for child endowments, then even low-income parents can achieve better outcomes without large private sacrifices. These policy-induced shifts in $E$ alter the perceived return to having more children or investing more per child.

\citet{becker1976child} also considered the possibility that endowment is not randomly assigned but may vary systematically with income. Higher-income households may produce children with higher $E$ due to better maternal nutrition, access to prenatal care, lower exposure to environmental risk, or assortative matching on traits associated with educational or occupational success. In these families, not only are the resources available for investment greater, but the potential gains from investment may also be higher, because children are better positioned to benefit from those inputs. This interaction deepens the divide between high- and low-income households, making it harder for disadvantaged families to catch up. As a result, inequality can persist or even widen across generations.

Finally, the model provides a mechanism for understanding how imperfections in credit markets can lead to persistent disadvantages. If parents with low income and low-$E$ children cannot borrow to finance quality-enhancing investment, then the next generation begins life with the same disadvantage. Without external intervention or structural change, this loop continues, which results in a pattern where poor families remain poor and rich families accumulate further advantage. The Becker-Tomes framework thus connects household-level decisions to bigger questions about the intergenerational transmission of human capital.

Parallel to Becker and Tomes’s static analysis of endowments, another strand of the literature developed a fully dynamic version of the Q-Q model by incorporating parental altruism toward children’s welfare. In this framework, introduced by \citet{barro1989fertility}, parents care not only about the number and quality of their children but also about the utility their descendants will enjoy in the future. Altruism in this context means that parents treat their children's utility as part of their own, thus extending the household’s objective across generations. For example, a parent may reduce personal consumption to pay for a child’s schooling, motivated not just by the child's immediate benefit but by the satisfaction the parent gains from the child’s long-term success. This leads to a formulation of dynastic utility, where the household's objective spans infinitely many periods and takes the form of a recursive altruistic structure. A representative formulation is

$$
U_{0} = \sum_{t=0}^{\infty} \beta^{t}\,u(c_{t}, n_{t}),
$$

where $c_t$ denotes the consumption of the $t$-th generation, $n_t$ the number of children, and $\beta \in (0,1)$ the intertemporal discount factor. Given this structure, having an extra child $n_t$ enters utility positively, but each child is assumed to receive the same utility as the parent if raised at a comparable standard of living. As a result, parents confront an intertemporal trade-off: having more children expands the number of future utility streams but also stretches current resources, since each child requires support. This trade-off gives rise to an Euler equation for optimal fertility choice, analogous to an optimal growth condition.

An implication of the dynastic model is that fertility decisions are sensitive to macroeconomic conditions, such as the interest rate or the rate of return on capital. A rise in interest rates increases the opportunity cost of channeling resources into children rather than saving, which tends to reduce current fertility—a substitution effect across generations. At the same time, higher returns make future generations wealthier, and this anticipated prosperity enters the utility calculations of parents in more complex ways. \citet{barro1989fertility} demonstrated that the model can account for observed fertility responses to economic fluctuations and policy interventions It can also explain historical phenomena such as postwar baby booms and subsequent fertility declines through shifts in returns or labor‐market opportunities.

In many dynastic models, child quality appears indirectly, often through the child’s future human capital or income. One variant assumes parents value the aggregate human‐capital stock of their children. This specification, combined with altruism, produces a similar trade‐off: concentrating resources in fewer children raises each child’s human capital, which raises the dynasty’s long‐run welfare. These intergenerational extensions link micro‐level fertility decisions to macroeconomic outcomes. By the late 1980s, work by Becker, Barro, and others had recast fertility as an endogenous choice that interacts with capital accumulation, income distribution, and policy. This laid the foundation for unified growth theories, which view the quantity–quality mechanism as central to demographic transition and long-run development.

### 2025-06-17 docs/paper.tex
write theoretical background (intergenerational q-q model)

While the early Q--Q models were static (one-period) representations,
subsequent contributions extended the framework to consider fertility
and child investment over multiple periods or even multiple generations.
A central development in this literature was the incorporation of
intergenerational human capital dynamics, where parents derive utility
not only from the number and quality of children in the present, but
also from the long-run outcomes of their offspring. These extensions
allowed child quality to evolve endogenously across time, rather than
being determined solely within a single period.

>>>>

While the early Q--Q models were static (one-period) representations,
subsequent contributions extended the framework to consider fertility
and child investment over multiple periods or even multiple generations.
The main development in this literature was the incorporation of
intergenerational human capital dynamics, where parents derive utility
not only from the number and quality of children in the present, but
also from the long-run outcomes of their offspring. These extensions
allowed child quality to evolve endogenously across time, rather than
being determined solely within a single period.

One of the earliest and most influential models of this kind was
proposed by \citet{becker1976child}. In their formulation, each child
enters the world with an exogenous endowment \(E\), which may reflect
factors such as cognitive ability (e.g., measured IQ or language
acquisition speed), early health status (e.g., birth weight or incidence
of neonatal complications), genetic predispositions (e.g., risk for
chronic illness, temperament, or neurodevelopmental traits), or family
background characteristics (e.g., parental education, household
stability, or neighborhood conditions). Furthermore, parents can augment
this endowment by investing resources \(q\) in the form of education,
nutrition, and other quality-enhancing inputs. The effective adult human
capital of the child might be expressed as \(H = E + f(q)\) (in a simple
additive form) or a multiplicative variant \(H = E \cdot f(q)\), where
\(f(q)\) is an increasing concave function of parental investment.

\citet{becker1976child} emphasized that variation in endowments \(E\)
can shape how parents allocate investments \(q\) across children. When
the productivity of investment increases with endowment, parents may
concentrate resources on children with higher \(E\), who are more likely
to convert additional investment into future success. In other cases,
parents may attempt to compensate for lower endowments by directing
greater investment toward disadvantaged children. Put simply,
child-specific variation in initial conditions affects not only outcomes
but also strategic parental choices. Because of this, the relationship
between income and the demand for child quality is not uniform. The
income elasticity of demand for quality may differ across households,
depending on the distribution of endowments within the family and across
the broader population.

Furthermore, \citet{becker1976child} showed that at low income levels,
much of what constitutes child quality comes from exogenous endowments,
i.e., factors like public education, neighborhood environment, or access
to basic healthcare that are not privately purchased. In these settings,
small increases in parental income may not lead to significant changes
in fertility or investment behavior. Since most of the child's future
outcomes are determined by the fixed endowment component, marginal
investment plays a smaller role. However, as income rises, private
resources become a larger part of what determines quality, and the
classic Q--Q trade-off begins to shape behavior. Parents begin to
allocate more income toward fewer children in order to enhance quality
through direct investment. Under certain theoretical conditions, such as
equal utility elasticities for quantity and quality, this framework
produces a non-monotonic relationship between income and fertility.
Fertility may decline as income rises at first, which reflects the
desire to invest more intensively per child. However, beyond some point,
once the marginal return to investment begins to flatten or saturate
relative to the fixed endowment, fertility may increase again.

The U-shaped prediction emerges only under specific assumptions, and its
validity depends on both the shape of the utility function and how
endowments relate to parental background. More broadly,
\citet{becker1976child} enriched the Q--Q framework by incorporating
elements that reflect real-world variation. They argued that not all
differences in child outcomes are the result of deliberate parental
choice. Random factors, biological traits, and socioeconomic settings
play a role. In this light, public policies, such as subsidized
schooling, early childhood programs, or universal healthcare, can
influence private fertility and investment decisions by shifting the
effective value of \(E\) across the population. If government programs
raise the floor for child endowments, then even low-income parents can
achieve better outcomes without large private sacrifices. These
policy-induced shifts in \(E\) alter the perceived return to having more
children or investing more per child.

\citet{becker1976child} also considered the possibility that endowment
is not randomly assigned but may vary systematically with income.
Higher-income households may produce children with higher \(E\) due to
better maternal nutrition, access to prenatal care, lower exposure to
environmental risk, or assortative matching on traits associated with
educational or occupational success. In these families, not only are the
resources available for investment greater, but the potential gains from
investment may also be higher, because children are better positioned to
benefit from those inputs. This interaction deepens the divide between
high- and low-income households, making it harder for disadvantaged
families to catch up. As a result, inequality can persist or even widen
across generations.

Finally, the model provides a mechanism for understanding how
imperfections in credit markets can lead to persistent disadvantages. If
parents with low income and low-\(E\) children cannot borrow to finance
quality-enhancing investment, then the next generation begins life with
the same disadvantage. Without external intervention or structural
change, this loop continues, which results in a pattern where poor
families remain poor and rich families accumulate further advantage. The
Becker-Tomes framework thus connects household-level decisions to bigger
questions about the intergenerational transmission of human capital.

Parallel to Becker and Tomes's static analysis of endowments, another
strand of the literature developed a fully dynamic version of the Q-Q
model by incorporating parental altruism toward children's welfare. In
this framework, introduced by \citet{barro1989fertility}, parents care
not only about the number and quality of their children but also about
the utility their descendants will enjoy in the future. Altruism in this
context means that parents treat their children's utility as part of
their own, thus extending the household's objective across generations.
For example, a parent may reduce personal consumption to pay for a
child's schooling, motivated not just by the child's immediate benefit
but by the satisfaction the parent gains from the child's long-term
success. This leads to a formulation of dynastic utility, where the
household's objective spans infinitely many periods and takes the form
of a recursive altruistic structure. A representative formulation is

\[
U_{0} = \sum_{t=0}^{\infty} \beta^{t}\,u(c_{t}, n_{t}),
\]

where \(c_t\) denotes the consumption of the \(t\)-th generation,
\(n_t\) the number of children, and \(\beta \in (0,1)\) the
intertemporal discount factor. Given this structure, having an extra
child \(n_t\) enters utility positively, but each child is assumed to
receive the same utility as the parent if raised at a comparable
standard of living. As a result, parents confront an intertemporal
trade-off: having more children expands the number of future utility
streams but also stretches current resources, since each child requires
support. This trade-off gives rise to an Euler equation for optimal
fertility choice, analogous to an optimal growth condition.

An implication of the dynastic model is that fertility decisions are
sensitive to macroeconomic conditions, such as the interest rate or the
rate of return on capital. A rise in interest rates increases the
opportunity cost of channeling resources into children rather than
saving, which tends to reduce current fertility---a substitution effect
across generations. At the same time, higher returns make future
generations wealthier, and this anticipated prosperity enters the
utility calculations of parents in more complex ways.
\citet{barro1989fertility} demonstrated that the model can account for
observed fertility responses to economic fluctuations and policy
interventions It can also explain historical phenomena such as postwar
baby booms and subsequent fertility declines through shifts in returns
or labor‐market opportunities.

In many dynastic models, child quality appears indirectly, often through
the child's future human capital or income. One variant assumes parents
value the aggregate human‐capital stock of their children. This
specification, combined with altruism, produces a similar trade‐off:
concentrating resources in fewer children raises each child's human
capital, which raises the dynasty's long‐run welfare. These
intergenerational extensions link micro‐level fertility decisions to
macroeconomic outcomes. By the late 1980s, work by Becker, Barro, and
others had recast fertility as an endogenous choice that interacts with
capital accumulation, income distribution, and policy. This laid the
foundation for unified growth theories, which view the quantity--quality
mechanism as central to demographic transition and long-run development.
